---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%"
)
```



# rcompendium



<!-- badges: start -->
[![R-CMD-check](https://github.com/ahasverus/rcompendium/workflows/R-CMD-check/badge.svg)](https://github.com/ahasverus/rcompendium/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/mFD)](https://CRAN.R-project.org/package=mFD)
[![License: GPL v2](https://img.shields.io/badge/License-GPL%20v2-blue.svg)](https://www.gnu.org/licenses/old-licenses/gpl-2.0.en.html)
[![LifeCycle](man/figures/lifecycle-experimental.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->


The goal of `rcompendium` is to create a research compendium, i.e. a predefined
folder structure with the following content:

```
compendium           # Project Root
├── .git
├── R/               # Contains R functions
├── man/             # Contains R functions helps (automatically updated)
├── analysis/
│   ├── data/        # Contains user raw data (.csv, .xlsx, .shp, etc.)
│   ├── rscripts/    # Contains R scripts to run analyses
│   ├── outputs/     # Contains outputs created by user
│   └── figures/     # Contains figures created by user
├── DESCRIPTION      # Project metadata (author, date, dependencies, etc.)
├── NAMESPACE
├── LICENSE          # License GPL >=2
├── LICENSE.md       # Content of the GPL >=2 License
├── make.R           # Master R scripts to run all analyses
├── README.md
├── README.Rmd
├── compendium.Rproj # RStudio project file (created by user)
├── .Rbuildignore    # Contains make.R and analysis/
└── .gitignore
```


**N.B. 1.** the research compendium can be built as an R package with 
`devtools::install()` but only R functions stored in the **R/** folder will
be available (do not forget to edit R functions documentation and to run
`devtools::document()`). 
The content of the **analysis/** folder must be run by sourcing the `make.R` 
file. All files created by user must be saved in the subfolders of **analysis/**.


**N.B. 2.** the files `DESCRIPTION` and `make.R` are written from templates and
are specific to myself.



## Installation



You can install the development version from [GitHub](https://github.com/) with:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("ahasverus/rcompendium")
```



## Usage

Under RStudio, you first need to create a **New project** (with or without
initializing git versioning). If you do not use RStudio IDE, you need to create
a **New folder** and to open R in this directory (by using `setwd()`).

**N.B.** the name of the compendium will be the same as the current directory.

Then you can create the compendium:

```{r eval=FALSE}
rcompendium::new_compendium()
```

You can now start working!



## Citation



Please cite this package as: 

> Casajus N. (2021) rcompendium: Create a Research Compendium. R package version 1.0.0.



## Code of Conduct



Please note that the `rcompendium` project is released with a 
[Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). 
By contributing to this project, you agree to abide by its terms.
