---
title: "Get started with rcompendium"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with rcompendium}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```



## Introduction



In the area of open science, making reproducible analyses is a strong prerequisite.
But sometimes it is difficult 1) to find the good structure for organizing files 
and 2) to setup the whole project.

The purpose of the package `rcompendium` is to make easier the creation of R
package/research compendium (i.e. a predefined files/folders structure) so that 
user can focus on the code/analysis instead of wasting time organizing files. 


<br />


A full ready-to-work structure will be set up with the following features: 

- Initialization of the [GIT](https://git-scm.com/) versioning (see below [GIT configuration]).
- Creation of a minimal R package structure (`DESCRIPTION` and `NAMESPACE` files, 
and `R/` and `man/` folders).
- Creation of additional files (`LICENSE.md`, `inst/CITATION`, etc.).
- Creation of a _Get started_ vignette in `vignettes/`.
- Setting the units tests process.
- Creation of a `README.Rmd` with HexSticker (template) and badges.
- Autocompletion of maintainer information (see below [Managing credentials]).
- Creation of a GitHub repository (see below [Creating a GitHub repo]).
- Configuration of GitHub Actions to automatically:
  - check and test package (`R CMD Check`);
  - report the code coverage (`covr`);
  - build and deploy website (`pkgdown`).


<br />


The strength of `rcompendium` is to create the whole structure in one command 
line by using the function `new_package()`. The default settings will produce
the following structure (where `pkg` is the name of the package):



```
pkg
│
├── pkg.Rproj                   # (optional) Created by user 
│
├── .git/                       # GIT tracking folder
├── .gitignore                  # Specific to R packages
|
├── .github/                    # (optional) GitHub Actions settings
│   └── workflows/
│       ├── pkgdown.yaml        # Configuration file to build & deploy website
│       ├── R-CMD-check.yaml    # Configuration file to check & test package
│       └── test-coverage.yaml  # Configuration file to build & deploy website
│
├── _pkgdown.yaml               # (optional) User website settings
│
├── R/                          # R functions
│   ├── fun-demo.R              # Example of an R function
│   └── pkg-package.R           # Dummy R file for package-level documentation
│
├── man/                        # R functions helps (automatically updated)
│   ├── print_msg.Rd            # Documentation of the demo R function
│   ├── pkg-package.Rd          # Package-level documentation
│   └── figures/                # Figures for the README 
│       └── hexsticker.png      # Template R package HexSticker
│
├── tests/
│   ├── testthat.R              # Units tests settings
│   └── testthat/               # Units tests folder
│       └── test-demo.R         # Units tests for the function print_msg()
|
├── vignettes/
│   └── pkg.Rmd                 # (optional) Package tutorial              [*]
│
├── DESCRIPTION                 # Project metadata                         [*]
├── NAMESPACE                   # Automatically generated
├── .Rbuildignore               # List of files/folders to be ignored while 
│                               # checking the package
├── inst/
│   └── CITATION                # BiBTeX entry to cite the R package       [*]
│
├── LICENSE                     # (optional) If License = MIT
├── LICENSE.md                  # Content of the chosen license
│
├── README.md                   # GitHub README (automatically generated)
└── README.Rmd                  # GitHub README (to knit)                  [*]


[*] These files are automatically edited but user needs to add manually 
    some information (title, description, version, etc.).
```

<br />

The second main function of `rcompendium` is `new_compendium()`. It will add the
following files/folders to the previous content:

```
...
├── analysis/                 # Proposed compendium
│   ├── data/                 # User raw data (.csv, .gpkg, etc.)
│   ├── rscripts/             # R scripts (no functions) to run analyses
│   ├── outputs/              # Outputs created by R scripts
│   ├── figures/              # Figures created by R scripts
│   └── paper/                # Article based on analyses
│
└── make.R                    # Master R scripts to run all analyses
```

The folder `analysis/` and the file `make.R` will be added to the `.Rbuildignore`
so that some parts of the compendium (essentially the R functions) can be installed
as an R package (with `devtools::install()`). Installing the package is optional.

<br/>

**WHAT'S INTERESTING ABOUT `rcompendium`?**

1. **Autocompletion of user credentials**. The package will automatically add 
some metadata (user information, package name, package version, etc.) to 
`DESCRIPTION`, `CITATION` and `README.Rmd` as the package.
[`usethis`](https://usethis.r-lib.org/) does.
2. **Detection of dependencies**. The function `add_dependencies()` will
check for all external dependencies (i.e. functions in external packages used in
the project) and automatically add these packages to the `DESCRIPTION` file (fields
`Imports` and `Suggests`). See `?add_dependencies` for further information.
3. **Minimal R version**. The function `add_r_depend()` can find the minimal 
R version required to run the project. This information can be added to the 
`DESCRIPTION` file as: `Depends: R (>= 9.9)`
4. **Automatic update**. The function `refresh()` will automatically update the 
project (re-build `.Rd` files, re-knit the `README.Rmd`, etc.). See [Usage] section below.
5. **GitHub Actions**. The package will setup CI/CD in background (optional).


<br/>


## Installation

You can install the development version from [GitHub](https://github.com/FRBCesab/rcompendium) 
with:

```{r eval=FALSE}
## Install < remotes > package (if not already installed) ----

if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}


## Install dev version of < rcompendium > from GitHub ----

remotes::install_github("FRBCesab/rcompendium")
```


<br/>


## Prerequisites

Before using the package `rcompendium` user must follow these three steps.



### GIT configuration

First ensure that **GIT** is correctly installed on your machine and linked to
RStudio. Read the Chapter 6 of 
[Happy Git and GitHub for the useR](https://happygitwithr.com/install-git.html).

You also need to store your GIT credentials locally (i.e. for the project) or 
globally (recommended). Run `gh::gh_whoami()` to see if your git and associated 
credentials are correctly configured. If so you should see something like:

```
{
"name": "John Doe",
"login": "jdoe",
"html_url": "https://github.com/jdoe",
(...)
}
```

Otherwise you might need to run:

```{r eval = FALSE}
gert::git_config_global(name = "user.name",   value = "John Doe")
gert::git_config_global(name = "user.email",  value = "john.doe@@domain.com")
gert::git_config_global(name = "github.user", value = "jdoe")
```

See `gert::git_config()` for further information.



### Creating a GitHub repo

To create the GitHub repository directly from R, the package `rcompendium` 
uses the function `usethis::use_github()`, an client to the GitHub REST API. 
The interaction with this API required an authentication method: a 
**GITHUB PAT** (Personal Access Token).

If you don't have a **GITHUB PAT** locally stored, you must:
1. Obtain a new one from your GitHub account. **Make sure to select 
at least the first two scopes (private repository and workflow)**
2. Store it in the `.Renviron` file by using `usethis::edit_r_environ()`
and adding the following line: `GITHUB_PAT='99z9...z9'`

Run `usethis::gh_token_help()` for more information about getting and 
configuring a **GITHUB PAT**.

If everything is well configured you should see something like after calling
`gh::gh_whoami()`:

```
{
"name": "John Doe",
"login": "jdoe",
"html_url": "https://github.com/jdoe",
"scopes": "delete_repo, repo, workflow",
"token": "99z9...z9"
}
```

And you will be able to create a GitHub repository directly from R!




### Managing Credentials

You can use the arguments `given`, `family`, `email`, and `orcid`
directly with the function `new_package()` (and others). But if you create 
a lot a projects (packages and/or compendiums) it can be frustrating in the 
long run.

An alternative is to use **ONCE AND FOR ALL** the function 
`set_credentials()` to permanently store this information in the 
`.Rprofile` file. If these arguments are set to `NULL` (default) each 
function of the package `rcompendium` will search in this `.Rprofile` file.
It will save your time (it's the purpose of this package).

Even if you have stored your information in the `.Rprofile` file you will 
always be able to modify them on-the-fly (i.e. by using arguments of the
`new_package()`) or permanently by re-running `set_credentials()`.



<br/>

## Usage


The recommended workflow is:

1. Create an empty RStudio project.
2. Store your credentials with `set_credentials()` (if not already done).
3. Run `new_package()` to create a new package structure (and the GitHub repository).
4. Edit metadata in `DESCRIPTION`, `CITATION`, and `README.Rmd`.
5. Implement, document & test functions (and analyses).
6. Update the project with `refresh()`. This will:
  - update `.Rd` files and `NAMESPACE`;
  - update external dependencies in `DESCRIPTION`;
  - update badges and re-knit `README.Rmd`;
  - check and test package integrity (`devtools::check()`);
  - run `make.R` if the project is a research compendium (optional because some
  analyses might be time consuming).
7. Repeat steps 5 and 6 while developing the package.

<br/>

In addition to these four setup functions (`set_credentials()`, `new_package()`, 
`new_compendium()`, and `refresh()`), the package `rcompendium` offers 28 other functions.
The functions can be grouped as:

- `add_*()`: functions to add/update/overwrite files, configuration settings, dependencies, badges, etc.;
- `get_*()`: functions to retrieve some information (available licenses, created functions, R version, etc.).

Visit the [Reference page](https://frbcesab.github.io/rcompendium/reference/index.html)
for further details.



<br/>



To sum up:

```{r eval = FALSE}
library(rcompendium)

## Define **ONCE FOR ALL** your credentials ----

set_credentials(given = "John", family = "Doe", 
                email = "john.doe@@domain.com", 
                orcid = "9999-9999-9999-9999", protocol = "ssh")

## Create an R package structure ----

new_package()


## Start developing R functions in R/ ----
## ...


## Update package (documentation, dependencies, README, check) ----

refresh()


## Update Repository Status (example) ----

add_repostatus_badge(status = "active")


## Change LICENSE (example) ----

get_licenses()
##                      tag                                            url
## 1                    MIT        https://choosealicense.com/licenses/mit
## 2                    CC0    https://choosealicense.com/licenses/cc0-1.0
## 3              CC BY 4.0  https://choosealicense.com/licenses/cc-by-4.0
## 4                  GPL-2    https://choosealicense.com/licenses/gpl-2.0
## 5                  GPL-3    https://choosealicense.com/licenses/gpl-3.0
## 6             GPL (>= 2)    https://choosealicense.com/licenses/gpl-2.0
## 7             GPL (>= 3)    https://choosealicense.com/licenses/gpl-3.0
## 8               LGPL-2.1   https://choosealicense.com/licenses/lgpl-2.1
## 9                 LGPL-3   https://choosealicense.com/licenses/lgpl-3.0
## 10         LGPL (>= 2.1)   https://choosealicense.com/licenses/lgpl-2.1
## 11           LGPL (>= 3)   https://choosealicense.com/licenses/lgpl-3.0
## 12                AGPL-3   https://choosealicense.com/licenses/agpl-3.0
## 13           AGPL (>= 3)   https://choosealicense.com/licenses/agpl-3.0
## 14 Apache License (== 2) https://choosealicense.com/licenses/apache-2.0
## 15 Apache License (>= 2) https://choosealicense.com/licenses/apache-2.0

add_license(license = "MIT")


## Update package (documentation, dependencies, README, check) ----

refresh()
```


And do not forget to commit changes and to push to GitHub to enjoy GitHub Actions!


<br/>


## Working with a compendium

Coming soon...


<br/>


## Contributing

You are welcome to contribute to the `rcompendium` project. Please read our 
[Contribution Guidelines](https://frbcesab.github.io/rcompendium/CONTRIBUTING.html).

Please note that the `rcompendium` project is released with a 
[Contributor Code of Conduct](https://frbcesab.github.io/rcompendium/CODE_OF_CONDUCT.html). 
By contributing to this project, you agree to abide by its terms.


<br/>


## Colophon

This package is the result of intense discussions and feedbacks from the training 
course [Data Toolbox for Reproducible Research in Computational Ecology](https://github.com/FRBCesab/datatoolbox) (in French). 

`rcompendium` was largely inspired by the package 
[`rrtools`](https://github.com/benmarwick/rrtools) developed by 
[Ben Marwick](https://github.com/benmarwick) and tries to respect the standard 
defined by the community. It re-implements several functions from the packages
[`devtools`](https://devtools.r-lib.org/) and [`usethis`](https://usethis.r-lib.org/).
**Special thanks to these developers!**
