---
title: "Get started with rcompendium"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with rcompendium}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```



## Introduction



In the area of open science, making reproducible analyses is a strong prerequisite.
But sometimes it is difficult 1) to find the good structure to organize files 
and 2) to set up the whole project. The aim of the package `rcompendium` is to 
make easier the creation of R package/research compendium (i.e. a predefined 
files/folders structure) so that users can focus on the code/analysis instead of 
wasting time organizing files.


<br />


A full ready-to-work structure will be set up with the following features: 

- Initialization of the [GIT](https://git-scm.com/) version control.
- Creation of a minimal R package structure (`DESCRIPTION` and `NAMESPACE` files, 
and `R/` and `man/` folders).
- Creation of additional files (`LICENSE.md`, `inst/CITATION`, etc.).
- Creation of a _Get started_ vignette in `vignettes/`.
- Setting the units tests process.
- Creation of a `README.Rmd` with HexSticker (template) and badges.
- Autocompletion of maintainer information.
- Creation of a GitHub repository.
- Configuration of GitHub Actions to automatically:
  - check and test package (`R CMD Check`);
  - report the code coverage (`covr`);
  - build and deploy website (`pkgdown`).


<br />


The strength of `rcompendium` is to create the whole structure in one command 
line by using the function `new_package()`. The default settings will produce
the following structure (where `pkg` is the name of the package):



```
pkg
│
├── pkg.Rproj                   # (optional) Created by user 
│
├── .git/                       # GIT tracking folder
├── .gitignore                  # Specific to R packages
|
├── .github/                    # (optional) GitHub Actions settings
│   └── workflows/
│       ├── pkgdown.yaml        # Configuration file to build & deploy website
│       ├── R-CMD-check.yaml    # Configuration file to check & test package
│       └── test-coverage.yaml  # Configuration file to build & deploy website
│
├── _pkgdown.yaml               # (optional) User website settings
│
├── R/                          # R functions
│   ├── fun-demo.R              # Example of an R function
│   └── pkg-package.R           # Dummy R file for package-level documentation
│
├── man/                        # R functions helps (automatically updated)
│   ├── print_msg.Rd            # Documentation of the demo R function
│   ├── pkg-package.Rd          # Package-level documentation
│   └── figures/                # Figures for the README 
│       └── hexsticker.png      # Template R package HexSticker
│
├── tests/
│   ├── testthat.R              # Units tests settings
│   └── testthat/               # Units tests folder
│       └── test-demo.R         # Units tests for the function print_msg()
|
├── vignettes/
│   └── pkg.Rmd                 # (optional) Package tutorial              [*]
│
├── DESCRIPTION                 # Project metadata                         [*]
├── NAMESPACE                   # Automatically generated
├── .Rbuildignore               # List of files/folders to be ignored while 
│                               # checking the package
├── inst/
│   └── CITATION                # BiBTeX entry to cite the R package       [*]
│
├── LICENSE                     # (optional) If License = MIT
├── LICENSE.md                  # Content of the chosen license
│
├── README.md                   # GitHub README (automatically generated)
└── README.Rmd                  # GitHub README (to knit)                  [*]


[*] These files are automatically edited but user needs to add manually 
    some information (title, description, version, etc.).
```

<br />

The second main function of `rcompendium` is `new_compendium()`. It will add the
following files/folders to the previous content:

```
...
├── data/                       # User raw data (.csv, .gpkg, etc.)        
├── rscripts/                   # R scripts (no functions) to run analyses 
├── outputs/                    # Outputs created by R scripts             
├── figures/                    # Figures created by R scripts             
└── paper/                      # Article based on analyses                
│
└── make.R                      # Master R scripts to run all analyses     
```

The folder `analysis/` and the file `make.R` will be added to the `.Rbuildignore`
so that some parts of the compendium (essentially the R functions) can be installed
as an R package (with `devtools::install()`). Installing the package is optional.



<br/>



## Installation



You can install the development version from [GitHub](https://github.com/FRBCesab/rcompendium) 
with:

```{r eval=FALSE}
## Install < remotes > package (if not already installed) ----
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

## Install dev version of < rcompendium > from GitHub ----
remotes::install_github("FRBCesab/rcompendium")
```



<br/>



## Prerequisites



Before using the package `rcompendium` you must follow these three steps.



### GIT configuration



First ensure that **GIT** is correctly installed on your machine and linked to
RStudio. Read the Chapter 6 of 
[Happy Git and GitHub for the useR](https://happygitwithr.com/install-git.html).

You also need to store your GIT credentials locally (i.e. for the project) or 
globally (recommended). Run `gh::gh_whoami()` to see if your git and associated 
credentials are correctly configured. If so you should see something like:

```
{
"name": "John Doe",
"login": "jdoe",
"html_url": "https://github.com/jdoe",
(...)
}
```

Otherwise you might need to run:

```{r eval = FALSE}
gert::git_config_global(name = "user.name",   value = "John Doe")
gert::git_config_global(name = "user.email",  value = "john.doe@@domain.com")
gert::git_config_global(name = "github.user", value = "jdoe")
```

See `gert::git_config()` for further information.



### Creating a GitHub repo



To create the GitHub repository directly from R, the package `rcompendium` 
uses the function `usethis::use_github()`, a client to the GitHub REST API. 
The interaction with this API requires an authentication method: a 
**GITHUB PAT** (Personal Access Token).

If you don't have a **GITHUB PAT** locally stored, you must:

1. Obtain a new one from your GitHub account. **Make sure to select 
at least the first two scopes (private repository and workflow)**
2. Store this token in the `.Renviron` file by using `usethis::edit_r_environ()`
and adding the following line: `GITHUB_PAT='99z9...z9'`

Run `usethis::gh_token_help()` for more information about getting and 
configuring a **GITHUB PAT**.

If everything is well configured, after calling `gh::gh_whoami()` you should see 
something like:

```
{
"name": "John Doe",
"login": "jdoe",
"html_url": "https://github.com/jdoe",
"scopes": "delete_repo, repo, workflow",
"token": "99z9...z9"
}
```

Then you will be able to create a GitHub repository directly from R!




### Managing credentials



You can use the arguments `given`, `family`, `email`, and `orcid`
directly with the function `new_package()` (and others). But an alternative is 
to use **ONCE AND FOR ALL** the function 
`set_credentials()` to permanently store this information in the 
`.Rprofile` file. If these arguments are set to `NULL` (default) each 
function of the package `rcompendium` will search in this `.Rprofile` file.
It will save you a lot of time (that is the purpose of this package).

Even if you have stored your information in the `.Rprofile` file you will 
always be able to modify them on-the-fly (i.e. by using arguments of the
`new_package()` and others) or permanently by re-running `set_credentials()`.



<br/>



## Usage


The recommended workflow is:

1. **IMPORTANT -** Create an empty RStudio project.
2. Store your credentials with `set_credentials()` (if not already done).
3. Run `new_package()` or `new_compendium()` to create a new package/research compendium structure (and the GitHub repository if required).
4. Edit metadata in `DESCRIPTION`, `CITATION`, and `README.Rmd`.
5. Implement, document & test functions.
6. Update the project with `refresh()`. This will:
  - update `.Rd` files and `NAMESPACE`;
  - update external dependencies in `DESCRIPTION`;
  - update badges and re-knit `README.Rmd`;
  - check and test package (`devtools::check()`);
  - run `make.R` if the project is a research compendium (optional because some
  analyses might be time consuming).
7. Repeat steps 5 and 6 while working on the project.

<br/>

In addition to these four setup functions (`set_credentials()`, `new_package()`, 
`new_compendium()`, and `refresh()`), the package `rcompendium` offers 29 other functions.
These can be grouped as:

- `add_*()`: functions to add/update/overwrite files, configuration settings, dependencies, badges, etc.;
- `get_*()`: functions to retrieve some information (available licenses, R version, etc.).

Visit the [Reference page](https://frbcesab.github.io/rcompendium/reference/index.html)
for further details.



<br/>



**To sum up:**

```{r eval = FALSE}
library(rcompendium)

## Define **ONCE FOR ALL** your credentials ----

set_credentials(given = "John", family = "Doe", 
                email = "john.doe@@domain.com", 
                orcid = "9999-9999-9999-9999", protocol = "ssh")

## Create an R package structure ----

new_package()


## Then... 
## ... implement and document R functions in R/
## ... write units tests in tests/testthat/


## Update package (documentation, dependencies, README, check) ----

refresh()


## Update Repository Status (example) ----

add_repostatus_badge(status = "active")


## Change LICENSE (example) ----

get_licenses()
add_license(license = "MIT")


## Update package (documentation, dependencies, README, check) ----

refresh()
```


And do not forget to commit changes and to push to GitHub to enjoy GitHub Actions!

**N.B.** Users can also use functions from the package [`usethis`](https://usethis.r-lib.org/)
to add some missing features (e.g. `data/` and package release tools).


<br/>


## Contributing

You are welcome to contribute to the `rcompendium` project. Please read our 
[Contribution Guidelines](https://frbcesab.github.io/rcompendium/CONTRIBUTING.html).

Please note that the `rcompendium` project is released with a 
[Contributor Code of Conduct](https://frbcesab.github.io/rcompendium/CODE_OF_CONDUCT.html). 
By contributing to this project, you agree to abide by its terms.


<br/>


## Colophon

This package is the result of intense discussions and feedback from the training 
course [Data Toolbox for Reproducible Research in Computational Ecology](https://github.com/FRBCesab/datatoolbox) (in French). 

`rcompendium` is largely inspired by the package 
[`rrtools`](https://github.com/benmarwick/rrtools) developed by 
[Ben Marwick _et al._](https://github.com/benmarwick) and tries to respect the standard 
defined by the community. **Special thanks to these developers!**
