---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(collapse  = TRUE,
                      comment   = "#>",
                      fig.path  = "man/figures/",
                      out.width = "100%")
```



# {{project_name}}


<!-- badges: start -->
<!-- badges: end -->


Research Compendium of the project **{{ PLEASE ADD A FEW WORDS }}**



### How to cite

Please cite this compendium as: 

> **{{ PLEASE ADD A CITATION }}**



### Content

This repository is structured as follow:

- [`data/`](https://github.com/{{github}}/{{project_name}}/tree/master/data):
contains all raw data required to perform analyses

- [`analyses/`](https://github.com/{{github}}/{{project_name}}/tree/{{branch}}/analyses/):
contains R scripts to run each step of the workflow

- [`outputs/`](https://github.com/{{github}}/{{project_name}}/tree/{{branch}}/outputs):
contains all the results created during the workflow

- [`figures/`](https://github.com/{{github}}/{{project_name}}/tree/{{branch}}/figures):
contains all the figures created during the workflow

- [`R/`](https://github.com/{{github}}/{{project_name}}/tree/{{branch}}/R):
contains R functions developed especially for this project

- [`man/`](https://github.com/{{github}}/{{project_name}}/tree/{{branch}}/man):
contains help files of R functions

- [`DESCRIPTION`](https://github.com/{{github}}/{{project_name}}/tree/{{branch}}/DESCRIPTION):
contains project metadata (author, date, dependencies, etc.)

- [`make.R`](https://github.com/{{github}}/{{project_name}}/tree/{{branch}}/make.R):
main R script to run the entire project by calling each R script stored in the `analyses/` folder



### Usage

Clone the repository, open R/RStudio and run:

```{r eval = FALSE}
source("make.R")
```



### Notes

- All required packages, listed in the `DESCRIPTION` file, will be installed (if necessary)
- All required packages and R functions will be loaded
- Some analyses listed in the `make.R` might take time
